---
title: "STATS501 Project EDA"
author: "Falak Shah, Shubham Pandey, Anuraag Ramesh"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
# loading libraries
library(tidyverse)
library(lubridate)
library(stringr)
library(ggpubr)
```

```{r}
# loading data
results = read.csv("results.csv")
goalscorers = read.csv("goalscorers.csv")
shootouts = read.csv("shootouts.csv")
```

```{r}
#preprocessing

df_raw <- read_csv("results.csv")

# assign matchID

matchID <- 1:nrow(df_raw)
df_raw <- data.frame(matchID, df_raw) 

df_join <- 
 df_raw %>% 
 select(home_team, away_team, matchID) %>% 
 gather(key = type, value = key_team, -matchID)

df_processed <- df_raw %>% mutate(year = year(date))

df_augmented <- 
  df_join %>% 
  left_join(df_processed) %>% 
  mutate(result = case_when(
    type == "home_team" & home_score > away_score ~ "win",
    type == "home_team" & home_score < away_score ~ "loss",
    type == "away_team" & away_score > home_score ~ "win",
    type == "away_team" & away_score < home_score ~ "loss",
    home_score == away_score ~ "draw"
  )) %>% 
  mutate(
    goals_for     = ifelse(type == "home_team", home_score, away_score),
    goals_against = ifelse(type == "home_team", away_score, home_score)
  ) %>% 
  mutate(
    tmp = 1
  ) %>% 
  spread(result, tmp, fill = 0)

df_summarized <- 
  df_augmented %>%
  group_by(year, key_team) %>% 
  summarize_at(vars(win, draw, loss, goals_for, goals_against), sum) 

write_csv(df_summarized, str_c(today(), "_summarized-yearly-results.csv"))
```


```{r}
df_summarized
```

# EDA

## Why negative binomial
```{r}
# distribution of all-time goals scored here (neg binom)
ggplot(data=df_augmented, aes(goals_for)) + 
  geom_histogram(binwidth = 1, 
                 col="blue", 
                 fill="red", 
                 alpha = .2) + 
  labs(title="Goals Scored by Team in Single Match") +
  labs(x="Number of Goals Scored in Match",y = "Count")

ggsave("whynb.png")
```

```{r}
mean(df_augmented$goals_for)
var(df_augmented$goals_for)
```

# Goals scored in a year

```{r}
p1 <- 
  ggplot(data=df_summarized, aes(goals_for)) + 
    geom_histogram(binwidth = 1,
                   col="blue", 
                   fill="red", 
                   alpha = .2) + 
    labs(title="Goals Scored by Team in Single Year") +
    labs(x="Number of Goals Scored in Year")
ggsave("goals_scored_year.png", plot = p1)
```


# Goals conceded in a year

```{r}
p2 <-
  ggplot(data=df_summarized, aes(goals_against)) + 
    geom_histogram(binwidth = 1,
                   col="blue", 
                  fill="red", 
                   alpha = .2) + 
    labs(title="Goals Allowed by Team in Single Year") +
    labs(x="Number of Goals Allowed in Year")
ggsave("goals_conc_year.png", plot = p2)
```

# Distribution of games across tournaments

```{r}
tournament = results %>% group_by(tournament) %>% summarise(n = n()) %>% arrange(desc(n))

p <- ggplot(tournament[c(1:10),], aes(x = reorder(tournament, n), y = n)) + geom_bar(stat = "identity",fill = "steelblue")
p + coord_flip() + theme_minimal() + xlab("Tournament") + ylab("Number of Games") + ggtitle("Number of games across different types of tournaments")
ggsave("tournaments.png")
```

# Number of games each year

```{r}
results$date = as.Date(results$date)
results$year = format(as.Date(results$date, format="%Y/%m/%d"),"%Y")
games = results %>% group_by(year) %>% summarise(n = n())

# Change the color
ggplot(data=games, aes(x=year, y=n)) +
  geom_line(color="red")+
  geom_point() + scale_x_discrete(breaks = seq(1872,2022,10)) + xlab("Year") +
  ylab("Number of Games") + ggtitle("Number of games from 1872 to 2022")

ggsave("number_of_games.png")

```

# Most attacking team

```{r}
df_summarized %>% group_by(key_team) %>% summarise(avg_scored = mean(goals_for)) %>% arrange(desc(avg_scored))
```

# Most defending team

```{r}
df_summarized %>% group_by(key_team) %>% summarise(avg_conc = mean(goals_against)) %>% arrange((avg_conc))
```

# Importance of Home vs Away

```{r}
df_summarized2 <- 
  df_augmented %>%
  group_by(year, key_team, type) %>% 
  summarize_at(vars(win, draw, loss, goals_for, goals_against), sum) %>% 
  mutate(total_matches = win + loss + draw) %>% 
  filter(year > 1910)

df_summarized2 = df_summarized2 %>% mutate(perc_win = (win/total_matches)*100,
                                           perc_loss = (loss/total_matches)*100)

df_summarized2 %>% arrange(desc(perc_win))

ggplot(df_summarized2, aes(x=type, y=perc_win, fill=type)) + 
    geom_boxplot() + xlab("Venue of Game") + ylab("Win Percentage") + ggtitle("Distribution of home vs away win percentage across years")

ggsave("home_vs_away.png")

df_summarized3 <- 
  df_augmented %>%
  group_by(year, type) %>% 
  summarize_at(vars(win, draw, loss, goals_for, goals_against), sum) %>% 
  mutate(total_matches = win + loss + draw) %>% 
  filter(year > 1910)

df_summ = df_summarized3 %>% mutate(win_perc = (win/total_matches)*100)
df_summ

ggplot(df_summ,aes(x= year, y= win_perc, group=type, color=type)) +
    geom_line() + xlab("Year") +
  ylab("Win Percentage") + ggtitle("Home vs Away win percentage across years")

ggsave("home_vs_away2.png")
```

# Importance of oppostion strenghts

# Most successful home teams

```{r}
df_summarized2$combined =  str_c(df_summarized2$key_team," ",df_summarized2$year)

home_df = df_summarized2 %>% filter(type == "home_team",total_matches >= 10) %>% arrange(desc(perc_win))
away_df = df_summarized2 %>% filter(type == "away_team",total_matches >= 10) %>% arrange(desc(perc_win))
home_df = home_df[1:10,]
away_df = away_df[1:10,]

# Most successful home teams
ggdotchart(home_df, x = "combined", y = "perc_win", # Custom color palette
           sorting = "descending",                       # Sort value in descending order
           add = "segments",
           color = "red",# Add segments from y = 0 to dots
           rotate = TRUE,
           palette = c("#00AFBB"),# Rotate vertically
                                           # Order by groups
           dot.size = 6,                                 # Large dot size
           label = round(home_df$perc_win),                        # Add mpg values as dot labels
           font.label = list(color = "white", size = 9, 
                             vjust = 0.5),               # Adjust label parameters
           ggtheme = theme_pubr(),
           xlab = "Country and Year",
           ylab = "Win Percentage",
           title = "10 Most Successful Home Teams"# ggplot2 theme
           )

ggsave("home_success.png")

```

# Least successful home teams

```{r}
# Least successful home teams
home_df = df_summarized2 %>% filter(type == "home_team",total_matches >= 10) %>% arrange(desc(perc_win))
home_df2 = tail(home_df,10)

ggdotchart(home_df2, x = "combined", y = "perc_win", # Custom color palette
           sorting = "descending",                       # Sort value in descending order
           add = "segments",
           color = "blue",# Add segments from y = 0 to dots
           rotate = TRUE,
           palette = c("#00AFBB"),# Rotate vertically
                                           # Order by groups
           dot.size = 6,                                 # Large dot size
           label = round(home_df2$perc_win),                        # Add mpg values as dot labels
           font.label = list(color = "white", size = 9, 
                             vjust = 0.5),               # Adjust label parameters
           ggtheme = theme_pubr(),
           xlab = "Country and Year",
           ylab = "Win Percentage",
           title = "10 Least Successful Home Teams"# ggplot2 theme
           )

ggsave("home_fail.png")
```

# Most successful away teams

```{r}
# Most successful away teams
away_df = df_summarized2 %>% filter(type == "away_team",total_matches >= 10) %>% arrange(desc(perc_win))
away_df2 = head(away_df,10)

ggdotchart(away_df2, x = "combined", y = "perc_win", # Custom color palette
           sorting = "descending",                       # Sort value in descending order
           add = "segments",
           color = "dark green",# Add segments from y = 0 to dots
           rotate = TRUE,
           palette = c("#00AFBB"),# Rotate vertically
                                           # Order by groups
           dot.size = 6,                                 # Large dot size
           label = round(away_df2$perc_win),                        # Add mpg values as dot labels
           font.label = list(color = "white", size = 9, 
                             vjust = 0.5),               # Adjust label parameters
           ggtheme = theme_pubr(),
           xlab = "Country and Year",
           ylab = "Win Percentage",
           title = "10 Most Successful Away Teams"# ggplot2 theme
           )
ggsave("away_success.png")
```

```{r}
results %>% filter(year >= 2000)
```

# Team strength metric according to generalized mixed effect model

```{r}
model = read.csv("quality_team.csv")
model = model[,c(1:4)]
model %>% arrange(desc(quality.metric))
```


```{r}
tc = c("Brazil","Spain","Argentina","France","England","Netherlands","Portugal","Germany","Colombia","Uruguay")
model_long = model %>% 
  pivot_longer(cols=c('attack', 'defense','quality.metric'),
                    names_to='type',
                    values_to='score')

test_model = model_long %>% filter(country %in% tc)

test_model$country <- factor(test_model$country, levels = c("Brazil","Spain","Argentina","France","England","Netherlands","Portugal","Germany","Colombia","Uruguay"))
test_model
```


```{r}
# Grouped
ggplot(test_model, aes(fill=type, y=score, x=country)) + 
    geom_bar(position="dodge", stat="identity") + coord_flip() + ylab("Overall Quality") + xlab("Country") +
  ggtitle("Ten countries with the highest quality (In increasing order)") +
  scale_fill_manual(values=c('Dark Green', 'Red', 'Light Blue'))

ggsave("model_plot.png")
```

# Yearly Team strength metric according to generalized mixed effect model

```{r}
plot_df = read.csv("quality_teamyear.csv")
plot_df = plot_df %>% mutate(overall_quality = overall_attack - overall_defence) %>% select(year,country,year_quality,overall_quality) %>% arrange(desc(year_quality))

test_df =  plot_df %>% group_by(country) %>% summarise(overall_quality)
distinct_df = distinct(test_df)

plot_df = plot_df %>% group_by(country) %>% summarise(max_qual = max(year_quality),min_qual = min(year_quality))
distinct_df
plot_df$overall = distinct_df$overall_quality
plot_df

plot_longer = plot_df %>% 
  pivot_longer(cols=c('max_qual', 'min_qual','overall'),
                    names_to='type',
                    values_to='score')
plot_longer

tp = c("Germany","Portugal","Netherlands")
plot_longer = plot_longer %>% filter(country %in% tp)
plot_longer
plot_longer$year = c("2010","2003","","2015","2002","","2022","2008","")
plot_longer

df <- plot_longer %>% mutate(
  country_year = paste(country,year, sep = " ")
)
df = df[,c(2:5)]

df = df[,c(4,1,2)]

# Grouped
ggplot(df, aes(fill=type, y=score, x=country_year)) + 
    geom_bar(position="dodge", stat="identity",width = 0.3) + coord_flip() + ylab("Overall Quality") + xlab("Country") +
  ggtitle("Country quality across years") +
  scale_fill_manual(values=c('Dark Green', 'Red', 'Light Blue'))

ggsave("model_plot2.png")

```

# Team strength metric according to bayesian mixed effect model

```{r}
quality_team = read.csv('quality_team_b.csv')
tc = c("Brazil","Spain","Argentina","France","England","Netherlands","Portugal","Germany","Colombia","Uruguay")
model_long = quality_team %>% 
  pivot_longer(cols=c('attack', 'defense','quality_metric'),
                    names_to='type',
                    values_to='score')

test_model = model_long %>% filter(country %in% tc)

test_model$country <- factor(test_model$country, levels = c("Brazil","England","Netherlands","Argentina","Spain","France","Germany","Portugal","Uruguay","Colombia"))
test_model %>% arrange(desc(score))

# Grouped
ggplot(test_model, aes(fill=type, y=score, x=country)) + 
    geom_bar(position="dodge", stat="identity") + coord_flip() + ylab("Overall Quality") + xlab("Country") +
  ggtitle("Ten countries with the highest quality (In increasing order)") +
  scale_fill_manual(values=c('Dark Green', 'Red', 'Light Blue'))

ggsave("bayes_plot1.png")
```

# Yearly Team strength metric according to bayesian mixed effect model

```{r}
quality_team_year = read.csv("quality_teamyear_b.csv")

plot_df = quality_team_year %>% mutate(overall_quality = overall_attack - overall_defence) %>% select(year,country,year_quality,overall_quality) %>% arrange(desc(year_quality))

test_df =  plot_df %>% group_by(country) %>% summarise(overall_quality)
distinct_df = distinct(test_df)

plot_df = plot_df %>% group_by(country) %>% summarise(max_qual = max(year_quality),min_qual = min(year_quality))
distinct_df
plot_df$overall = distinct_df$overall_quality
plot_df

plot_longer = plot_df %>% 
  pivot_longer(cols=c('max_qual', 'min_qual','overall'),
                    names_to='type',
                    values_to='score')
plot_longer

tp = c("Germany","Portugal","Netherlands")
plot_longer = plot_longer %>% filter(country %in% tp)
plot_longer
plot_longer$year = c("2010","2003","","2015","2002","","2022","2008","")
plot_longer

df <- plot_longer %>% mutate(
  country_year = paste(country,year, sep = " ")
)
df = df[,c(2:5)]

df = df[,c(4,1,2)]

# Grouped
ggplot(df, aes(fill=type, y=score, x=country_year)) + 
    geom_bar(position="dodge", stat="identity",width = 0.3) + coord_flip() + ylab("Overall Quality") + xlab("Country") +
  ggtitle("Country quality across years") +
  scale_fill_manual(values=c('Dark Green', 'Red', 'Light Blue'))

ggsave("bayes_plot2.png")

```

