---
title: "STATS501 Project Modelling"
author: "Anuraag Ramesh, Falak Shah, Shubham Pandey"
date: "Apr 23, 2023"
output:
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE, comment=NA, out_width=100, warning = FALSE, message = FALSE)
library(tidyverse)
library(lubridate)
library(lme4)
library(brms)
library(rstanarm)
library(rstan)
```

Initial Data Processing using results.csv (Match results from 1872 to present)

```{r}
#Initial data
df_initial <- read_csv("results.csv")
#We assign a match ID to make each record unique
matchID <- 1:nrow(df_initial)
df_initial <- data.frame(matchID, df_initial)

#Add a Year column
df_year <- 
  df_initial %>% 
  mutate(year = year(date))
```

```{r}
#Top Goalscorers for a country in a year
goal_scorers = read_csv('goalscorers.csv')
goal_scorers$year = year(goal_scorers$date)

temp = goal_scorers %>% group_by(year, team, scorer) %>% count() 
new = temp %>% group_by(year, team) %>% filter(n == max(n)) %>% filter(row_number()==1)
head(new %>% arrange(desc(n)), 20)
```

Harry Kane in 2021 scored 16 goals for England, which implies an amazing year for the country.    

```{r}
# Subset the dataset to required columns
df_sub <- 
  df_initial %>% 
  select(home_team, away_team, matchID) %>% 
  gather(key = type, value = key_team, -matchID)

# We add a home_team and away_team score in the dataset using one-hot encoding
df_unique <- 
  df_sub %>% 
  left_join(df_year) %>% 
  mutate(result = case_when(
    type == "home_team" & home_score > away_score ~ "win",
    type == "home_team" & home_score < away_score ~ "loss",
    type == "away_team" & away_score > home_score ~ "win",
    type == "away_team" & away_score < home_score ~ "loss",
    home_score == away_score ~ "draw")) %>% 
  mutate(
    goals_for     = ifelse(type == "home_team", home_score, away_score),
    goals_against = ifelse(type == "home_team", away_score, home_score)) %>% 
  mutate(
    tmp = 1) %>% spread(result, tmp, fill = 0)
```


```{r}
# We consider data after 1910s, because there wasn't a lot of countries 
# playing football before this
df_homeaway <- 
  df_unique %>%
  group_by(year, key_team, type) %>% 
  summarize_at(vars(win, draw, loss, goals_for, goals_against), sum) %>% 
  mutate(total_matches = win + loss + draw) %>% 
  filter(year > 1910)

# We add a win percentage column
df_homeaway = df_homeaway %>% mutate(win_perc = win/total_matches)

#Preparing the dataset for the mixed model
df_mixed <- 
  df_unique %>%
  group_by(year, key_team) %>% 
  summarize_at(vars(win, draw, loss, goals_for, goals_against), sum) %>% 
  mutate(total_matches = win + loss + draw) %>% 
  filter(year > 1910)

df_mixed = df_mixed %>% mutate(win_perc = win/total_matches)
```


```{r}
# Reducing to only required variables to create the model
df_mixed1 <- 
  df_unique %>% 
  mutate(
    team = ifelse(type == "home_team", home_team, away_team),
    goals = ifelse(type == "home_team", home_score, away_score),
    opponent = ifelse(type == "home_team", away_team, home_team),
    type = ifelse(neutral == TRUE, "neutral", type)
  ) %>% 
  select(year, team, goals, opponent, type, date) %>% 
  filter(year > 1910)

# Select and change the name of variables to make sense
df_final = df_mixed1 %>% left_join(df_mixed,by=c('year'='year', 
                                    'opponent'='key_team')) %>%
  select(year, date, team, goals, opponent, type, opp_win = win_perc)
```


```{r}
# We can filter to all matches after 1970s to reduce running time for the model 
#and remove the test set(which is the 2022 world cup)
df_finalsub = df_final %>% filter(year >= 1970 & year < 2023)
df_finalsub = df_finalsub %>% filter(date < as.Date('2022-11-01'))
nrow(df_finalsub)
```

We have 36733 different matches across the past 50 years from the 1970s for training the model.

### Mixed Models

After pre-processing the data, we create a mixed effects model using team, opponent, team per year and opponent per year as random effects, with opponent win percentage as the fixed effect.  

```{r}
# First mixed model
#Takes 14 hours to run 
mixed_model <- glmer.nb(goals ~ opp_win + (1 | team/year) + (1| opponent/year) + 
                    (1 | type), df_finalsub)

saveRDS(mixed_model, 'mixed_model1.RDS')
```

```{r}
#Import the RDS file if model not run
mixed_model = read_rds('mixed_model.RDS')

mixed_model_out <- predict(mixed_model, type = "response")

#Plot for predicted vs actual
plot(mixed_model_out, df_finalsub$goals,
     main='Predicted Goals vs Actual Goals scored',
     xlab='Predicted', ylab='Actual')

# Plot a regression line
abline(lm(df_finalsub$goals ~ mixed_model_out), col = 2, lwd = 3)
```

```{r}
# Residual plots for mixed model
plot(mixed_model_out, residuals(mixed_model),
     main='Residuals vs Fitted',
     xlab='Fitted', ylab='Residuals')
```

```{r}
#Quality metrics plot functions

quality_plot <- function(model_coef){
  a = data.frame(model_coef$opponent)
  b = data.frame(model_coef$team)
  a$country = rownames(a)
  b$country = rownames(b)
  new = a %>% left_join(b, by = 'country') %>%
    select(country, attack = X.Intercept..y, defense = X.Intercept..x)
  new = new %>% mutate('quality_metric' = attack - defense)
  print(head(new), 10)
  return(new)
}

quality_plotyear <- function(model_coef){
  a = data.frame(model_coef$opponent)
  b = data.frame(model_coef$team)
  c = data.frame(model_coef$`year:opponent`)
  d = data.frame(model_coef$`year:team`)
  a$country = rownames(a)
  b$country = rownames(b)
  c$val = rownames(c)
  d$val = rownames(d)
  c = c %>% separate(val, c('year', 'country'), sep = ':')
  d = d %>% separate(val, c('year', 'country'), sep = ':')
  new1 = c %>% left_join(d, by = c('year', 'country')) %>%
    left_join(a, by = 'country') %>% left_join(b, by = 'country') %>%
    select(year, country, year_defence = X.Intercept..x, year_attack = X.Intercept..y, 
           overall_attack = X.Intercept..y.y,
           overall_defence = X.Intercept..x.x)

  new1 = new1 %>% mutate(year_quality = (overall_attack + year_attack) - (overall_defence + year_defence))
  
  new1 %>% arrange(desc(year_quality))
  print(head(new1), 10)
  return(new1)
}
```

```{r}
#Create for the frequentist mixed model
mixed_model_coef <- coef(mixed_model)
q1 = quality_plot(mixed_model_coef)
write_csv(q1, 'quality_team.csv')
q2 = quality_plotyear(mixed_model_coef)
write_csv(q2, 'quality_teamyear.csv')
```

```{r}
#Creating the test dataset for the world cup to test our models
df_test = df_final %>% filter(date >= as.Date('2022-11-20') & (date < as.Date('2022-12-19')))
df_test = df_test %>% select(-goals)
```

### Bayesian Mixed Models

Now, let us move on to the bayesian methods, we use the `stanarm` package. 

```{r}
#We create a beta prior to test
prior = prior("beta", 1, 1)

# We just use a vague flat prior instead for our model with 4 chains
bayesian_mixed = stan_glmer.nb(goals ~ opp_win + (1 | team/year) + (1| opponent/year) + (1 | type), 
              data = df_finalsub, cores = 4, chains = 4, iter = 1000)

saveRDS(bayesian_mixed, 'bayesian_model.RDS')
```

```{r}
#Import the RDS file if model not run
bayesian_mixed = read_rds('bayesian_model.RDS')

bayesian_coef = coef(bayesian_mixed)

#MCMC diagnostics

#First we check the R hat values and see if they are lesser than 1.1
any(summary(bayesian_mixed)[, "Rhat"] > 1.1)

#Then we check if the effective sample size (ESS) is lower than 1 in a lot of cases
hist(summary(bayesian_mixed)[, "n_eff"] / 1000)
```

The model seems to have converged based on Rhat and ESS.

```{r}
#Bayesian plots to compare
#Create for the bayesian mixed model
q3 = quality_plot(bayesian_coef)
write_csv(q3, 'quality_team_b.csv')
q4 = quality_plotyear(bayesian_coef)
write_csv(q4, 'quality_teamyear_b.csv')
```


### Testing phase for the world cup

```{r}
#Prediction for frequentist mixed method
df_test1 = copy(df_test)
df_test1 = df_test1 %>% filter(!team %in% c('Taiwan', 'Russia') & 
                      !opponent %in% c('Taiwan', 'Russia'))
mixed_goals_p = predict(mixed_model, df_test1)

df_test1$goals = mixed_goals_p
```

```{r}
#Interpretation
val = nrow(df_test1)/2
temp = df_test1 %>% slice(1:val)
temp1 = df_test1 %>% slice(val + 1:nrow(df_test1)) %>% select(goals2 = goals)
match_pred = cbind(temp, temp1)

#Only considering matches in the World Cup
match_pred = match_pred %>% filter(type == 'neutral' | team == 'Qatar' | opponent == 'Qatar') %>% slice(5: nrow(match_pred))
```

```{r}
#Only considering Group Stage games for fair comparison
match_pred = match_pred %>% slice(1:48)

final_winner = match_pred %>% 
    mutate(
      team_winner = case_when(
    goals > goals2  ~ team,
    goals < goals2  ~ opponent,
    goals == goals2  ~ "draw")
    )

#Predicting the teams to qualify
final_winner %>% group_by(team_winner) %>% count() %>% arrange(desc(n))
```
Frequentist model predictions. 

Actual: Netherlands, Senegal, England, United States, Argentina, Poland, France, Australia, Japan, Spain, Morocco, Croatia, Brazil, Switzerland, South Korea, Portugal

Predicted: Netherlands, Ecuador, England, Iran, Argentina, Poland, France, Denmark, Belgium, Croatia, Spain, Germany, Brazil, Serbia, Portugal, Uruguay

Frequentist Mixed Effects Model : 9/16  = 56.25%

```{r}
#Prediction for bayesian mixed method
bayes_goals_p = posterior_predict(bayesian_mixed, df_test, 2000)
df_test2 = copy(df_test)
goals_b = round(colMeans(bayes_goals_p))
df_test2$goals = goals_b
```

```{r}
#Interpretation
val = nrow(df_test2)/2
temp = df_test2 %>% slice(1:val)
temp1 = df_test2 %>% slice(val + 1:nrow(df_test1)) %>% select(goals2 = goals)
match_pred = cbind(temp, temp1)

match_pred = match_pred %>% filter(type == 'neutral' | team == 'Qatar' | opponent == 'Qatar') %>% slice(5: nrow(match_pred))

#Only considering Group Stage games for fair comparison
match_pred = match_pred %>% slice(1:48)

val = nrow(df_test2)/2
temp = df_test2 %>% slice(1:val) %>% select(goals2 = goals)
temp1 = df_test2 %>% slice(val + 1:nrow(df_test1)) 
match_pred1 = cbind(temp1, temp)

match_pred1 = match_pred1 %>% filter(type == 'neutral' | team == 'Qatar' | opponent == 'Qatar') %>% slice(5: nrow(match_pred1))

#Only considering Group Stage games for fair comparison
match_pred1 = match_pred1 %>% slice(1:48)

match_pred3 = rbind(match_pred, match_pred1)
```

```{r}
#Predicting the teams to qualify
final_winner = match_pred3 %>% 
  mutate(value = case_when(
    goals > goals2 ~ "win",
    goals < goals2 ~ "loss",
    goals == goals2 ~ "draw")) %>%
  mutate(
    tmp = 1) %>% spread(value, tmp, fill = 0)

#Finding the top teams
final_winner1 = final_winner %>% 
  group_by(team) %>% 
  summarize_at(vars(win, draw, loss), sum) %>%
  mutate(points = (win * 3 + draw)) %>%
  arrange(desc(points))
```

```{r}
#All 8 groups predicted tables
group1 = final_winner1 %>% filter(team %in% c('Netherlands', 'Senegal', 'Ecuador', 'Qatar'))
print(group1)

group2 = final_winner1 %>% filter(team %in% c('England', 'United States', 'Iran', 'Wales'))
print(group2)

group3 = final_winner1 %>% filter(team %in% c('Argentina' , 'Poland', 'Mexico', 'Saudi Arabia'))
print(group3)

group4 = final_winner1 %>% filter(team %in% c('France', 'Denmark', 'Australia', 'Tunisia'))
print(group4)

group5 = final_winner1 %>% filter(team %in% c('Japan', 'Spain', 'Germany', 'Costa Rica'))
print(group5)

group6 = final_winner1 %>% filter(team %in% c('Morocco', 'Croatia', 'Belgium', 'Canada'))
print(group6)

group7 = final_winner1 %>% filter(team %in% c('Brazil', 'Switzerland', 'Serbia', 'Cameroon'))
print(group7)

group8 = final_winner1 %>% filter(team %in% c('South Korea', 'Portugal', 'Ghana', 'Uruguay'))
print(group8)
```

Bayesian model predictions

Actual: Netherlands, Senegal, England, United States, Argentina, Poland, France, Australia, Japan, Spain, Morocco, Croatia, Brazil, Switzerland, South Korea, Portugal

Predicted: Netherlands, Senegal, England, United States, Argentina, Poland, France, Australia, Spain, Germany, Croatia, Belgium,  Brazil, Serbia, Portugal, Uruguay

Bayesian Mixed Effects Model : 12/16  = 75%
